Types for R, Empirically Designed and Evaluated

### 2020-05-07 paper/main.tex
Reformat text and add citations

contractr works by modifying function definitions to insert calls to our type-checking function. When the modified function is called, our type-checking function gets executed first.  The type-checker is invoked on each call to the function on all of its arguments and return value. There are two interesting challenges that need to be tackled to make this work reliably, without surprises for the end-users. Firstly, we need to correctly type-check arguments. R is a non-strict language (citation to my lazy paper); when a function is called, instead of getting values as arguments which can be immediately type-checked, we get unevaluated code thunks. We can force the evaluation of these thunks to obtain values for type-checking but that will not preserve the language’s semantics. This will lead to incorrect results and surprises for the end-users. To preserve the original semantics, we modify the thunks to call our type-checker on the value obtained when they are forced, along with the context necessary to associate the thunk to the function and its parameter (for failure messages). This works, but with one wrinkle. GNU R has a bytecode compiler that can sometimes optimize away thunks, in those cases, we don’t get thunks as arguments but values that can be immediately type-checked. However, we still can’t type-check the value and issue failure messages because that would violate the semantics. Non-strict semantics require that any computation related to the argument happens on the first use of the argument. If this argument was never used, it would be incorrect to type-check it. To make this work correctly, we explicitly wrap the argument value in a thunk with a call to the type-checker as in the previous case.  Secondly, we need to correctly type-check function return values. In R, the result of function evaluation is the result of evaluating the last expression in its body; so a return call is absent from most function definitions. So there are many potential subexpressions which need to be type-checked aginst the return type. For this, we register our type-checker to be called on the return value in a function exit hook. This hook is executed in the function’s environment after it has finished executing. While this works, we also have to contend with yet another wrinkle, R interpreter can perform longjumps in its C implementation, which causes R functions on the stack to be discarded. When they are discarded, their exit hooks are called, which in our case calls the type-checker. But, these functions are in the middle of an active computation, so they don’t have a return value to type-check. We deal with this problem by allocating a unique object which serves as the return value for calls that are discarded. Our exit hook does not call the type-checker if this unique value happens to be the return value of the function.

Type declarations can be made available to contractr in three ways: as part of its internal database of types (which already has type signatures for our corpus of 500 packages), as part of a designated TYPEDECLARATION file supplied by the package author and installed along with the package, or through a user-level API function insert\_contract that lets the user insert contract to a custom function with the type declaration provided as a string argument.
Through this package, we intend to provide a clear path for the R developers to add types to untyped package code. Hence, we have paid a lot of attention to usability in its design. 
Firstly, the package enables type-checking for all packages with type signatures, automatically, on importing, offering a seamless and hassle-free experience for both package developers and end-users. A mere library(contractr) is enough to insert contracts, both, in packages pre-loaded in the user’s workspace and in packages that will be loaded eventually. For this, contractr relies heavily on R’s reflective and dynamic capabilities. Upon loading contractr scans all the pre-loaded packages in the user’s workspace and inserts contracts in functions for which type signatures are available. For all other packages installed on user’s machine but not yet loaded in the workspace, contractr sets up package load hooks. The hooks get executed when these packages get loaded and insert contracts to the package functions. The ensures that users do not fiddle manually with the contractr API to enable type-checking for packages loaded eventually. Furthermore, contractr automatically removes contracts from all the package functions and restores them to their original state when it is unloaded, an operation that is rarely performed but is possible, nonetheless.
Secondly, we enable package authors to supply type declarations to contractr for their package functions, without requiring any modification to package code. This is accomplished by providing type declarations inside roxygen2 function comment blocks. roxygen2 is an R package that enables authors to add documentation to R functions in the form of organized plaintext sections which are automatically exported to R’s latex style custom documentation format. roxygen2 is used by … R packages in CRAN. It also provides an extension API for processing metadata in custom documentation tag sections. A package can register a method to be called when this tag is encountered. The metadata for that tag section is made available to the registrant before package build and installation steps. contractr uses this API to implement a @type tag. Our roxygen2 hook extracts type declarations from these sections from all the documented functions of the package and stores then in a TYPEDECLARATION file inside the package folder. This files is copied as is on package installation and automatically picked up contractr when the two packges are loaded.
Here is an example of how a roxygen code block looks like with our type declaration:

@@

In order to ensure that our type system design aligns with the day-to-day usage of R, we performed a corpus analysis of some of the most widely used R packages.
We built our dynamic analysis in the R-dyntrace \AT{cite something?} dynamic analysis framework, which is itself built on a modified version of R targeting R version 3.5.0---we refer the reader to Section~\ref{sec:r-dyntrace} for more details on R-dyntrace.

>>>>

contractr works by modifying function definitions to insert calls to our
type-checking function. When the modified function is called, our type-checking
function gets executed first.  The type-checker is invoked on each call to the
function on all of its arguments and return value.

%
%
\subsubsection{Challenges}
There are two interesting challenges that need to be tackled to make this work
reliably, without surprises for the end-users. Firstly, we need to correctly
type-check arguments. R is a non-strict language~\cite{oopsla19};
when a function is called, instead of getting values as arguments which can be
immediately type-checked, we get unevaluated code thunks. We can force the
evaluation of these thunks to obtain values for type-checking but that will not
preserve the language’s semantics. This will lead to incorrect results and
surprises for the end-users. To preserve the original semantics, we modify the
thunks to call our type-checker on the value obtained when they are forced,
along with the context necessary to associate the thunk to the function and its
parameter (for failure messages). This works, but with one wrinkle. GNU R has a
bytecode compiler that can sometimes optimize away thunks, in those cases, we
don’t get thunks as arguments but values that can be immediately type-checked.
However, we still can’t type-check the value and issue failure messages because
that would violate the semantics. Non-strict semantics require that any
computation related to the argument happens on the first use of the argument. If
this argument was never used, it would be incorrect to type-check it. To make
this work correctly, we explicitly wrap the argument value in a thunk with a
call to the type-checker as in the previous case.  Secondly, we need to
correctly type-check function return values. In R, the result of function
evaluation is the result of evaluating the last expression in its body; so a
return call is absent from most function definitions. So there are many
potential subexpressions which need to be type-checked aginst the return type.
For this, we register our type-checker to be called on the return value in a
function exit hook. This hook is executed in the function’s environment after it
has finished executing. While this works, we also have to contend with yet
another wrinkle, R interpreter can perform longjumps in its C implementation,
which causes R functions on the stack to be discarded. When they are discarded,
their exit hooks are called, which in our case calls the type-checker. But,
these functions are in the middle of an active computation, so they don’t have a
return value to type-check. We deal with this problem by allocating a unique
object which serves as the return value for calls that are discarded. Our exit
hook does not call the type-checker if this unique value happens to be the
return value of the function.

%
%
\subsubsection{Type Declarations}
Type declarations can be made available to contractr in three ways: as part of
its internal database of types (which already has type signatures for our corpus
of 500 packages), as part of a designated TYPEDECLARATION file supplied by the
package author and installed along with the package, or through a user-level API
function insert\_contract that lets the user insert contract to a custom
function with the type declaration provided as a string argument.

%
%
\subsubsection{Usability}
Through this package, we intend to provide a clear path for the R developers to
add types to untyped package code. Hence, we have paid a lot of attention to
usability in its design.
Firstly, the package enables type-checking for all packages with type
signatures, automatically, on importing, offering a seamless and hassle-free
experience for both package developers and end-users. A mere library(contractr)
is enough to insert contracts, both, in packages pre-loaded in the user’s
workspace and in packages that will be loaded eventually. For this, contractr
relies heavily on R’s reflective and dynamic capabilities. Upon loading
contractr scans all the pre-loaded packages in the user’s workspace and inserts
contracts in functions for which type signatures are available. For all other
packages installed on user’s machine but not yet loaded in the workspace,
contractr sets up package load hooks. The hooks get executed when these packages
get loaded and insert contracts to the package functions. The ensures that users
do not fiddle manually with the contractr API to enable type-checking for
packages loaded eventually. Furthermore, contractr automatically removes
contracts from all the package functions and restores them to their original
state when it is unloaded, an operation that is rarely performed but is
possible, nonetheless.
Secondly, we enable package authors to supply type declarations to contractr for
their package functions, without requiring any modification to package code.
This is accomplished by providing type declarations inside roxygen2 function
comment blocks. roxygen2 is an R package that enables authors to add
documentation to R functions in the form of organized plaintext sections which
are automatically exported to R’s latex style custom documentation format.
roxygen2 is used by … R packages in CRAN. It also provides an extension API for
processing metadata in custom documentation tag sections. A package can register
a method to be called when this tag is encountered. The metadata for that tag
section is made available to the registrant before package build and
installation steps. contractr uses this API to implement a @type tag. Our
roxygen2 hook extracts type declarations from these sections from all the
documented functions of the package and stores then in a TYPEDECLARATION file
inside the package folder. This files is copied as is on package installation
and automatically picked up contractr when the two packges are loaded.
Here is an example of how a roxygen code block looks like with our type declaration:

@@

In order to ensure that our type system design aligns with the day-to-day usage of R, we performed a corpus analysis of some of the most widely used R packages.
We built our dynamic analysis in the R-dyntrace~\cite{oopsla19} dynamic analysis framework, which is itself built on a modified version of R targeting R version 3.5.0---we refer the reader to Section~\ref{sec:r-dyntrace} for more details on R-dyntrace.
